\chapter{Requirements}
\label{chap:requirements}

In this chapter, we detail both functional and non-functional requirements that guide the design and implementation of the proposed system. These requirements are organized around four key components, each pairing a model with an associated algorithm:

\begin{enumerate}
    \item A model and an algorithm for multi-dimensional classification of vulnerabilities,
    \item A model and an algorithm to compute a classification for a known vulnerability in the context of a specific software environment,
    \item A model and an algorithm to provide vulnerability remediation recommendations,
    \item A model and an algorithm to rank-order all known classified vulnerabilities.
\end{enumerate}

\section{Functional Requirements}
This section outlines the functionalities the system must provide to meet the core objectives of vulnerability classification, scoring, remediation, and prioritization.

\subsection{RQ1: A Model and an Algorithm for Multi-Dimensional Classification of Vulnerabilities}
\label{req:RQ1}
To handle vulnerabilities effectively, the system requires a model and an algorithm that address multiple classification dimensions and ensure user-friendly transparency in scoring.

\begin{itemize}
    \item \textbf{RQ1.1: Multiple Classification Dimensions}\\
    \label{req:RQ1.1}  
    The system must integrate multiple dimensions to classify vulnerabilities.

    \item \textbf{RQ1.2: Handling of Missing Data}\\
    \label{req:RQ1.2}  
    The system must be able to manage missing or incomplete data while still providing meaningful classification results.

    \item \textbf{RQ1.3: User Interface for Score Explanation}\\
    \label{req:RQ1.3}
    The system must present classification details in an intuitive and transparent manner.
    \begin{itemize}
        \item \textbf{RQ1.3.1: Score-Details Button}\\
        \label{req:RQ1.3.1}  
        Include a “Show Score-Details” button that displays key classification details.

        \item \textbf{RQ1.3.2: Tooltip Explanations}\\
        \label{req:RQ1.3.2}  
        Provide hover-based tooltips for each dimension (e.g., CVSS vector attributes) to explain how each attribute affects severity.

        \item \textbf{RQ1.3.3: Prefilled \ac{CVSS} Calculator Link}\\
        \label{req:RQ1.3.3}  
        Offer a direct link to a prefilled \ac{CVSS} calculator, enabling users to quickly explore alternative scoring scenarios.
    \end{itemize}
\end{itemize}

\subsection{RQ2: A Model and an Algorithm to Compute a Classification for a Known Vulnerability}
\label{req:RQ2}
This set of requirements focuses on computing a classification score for any known vulnerability, taking into account the specific software context.

\begin{itemize}
    \item \textbf{RQ2.1: Data Gathering}\\
    \label{req:RQ2.1}  
    The system must collect vulnerability data from external sources.

    \item \textbf{RQ2.2: Vulnerability Categorization}\\
    \label{req:RQ2.2}  
    The system must classify vulnerabilities (e.g., Critical, High, Medium, Low) based on the computed severity scores.
\end{itemize}

\subsection{RQ3: A Model and an Algorithm to Make Recommendations for Remediation}
\label{req:RQ3}
Beyond classification, the system should offer actionable guidance to address discovered vulnerabilities.

\begin{itemize}
    \item \textbf{RQ3.1: Remediation Guidance}\\
    \label{req:RQ3.1}  
    The system must generate targeted recommendations for remediating identified vulnerabilities.

    \item \textbf{RQ3.2: Role-Based Recommendations}\\
    \label{req:RQ3.2}  
    Recommendations must be tailored to different stakeholder roles (e.g., developers, security coordinators) and asset criticality.

    \item \textbf{RQ3.3: Action Categorization}\\
    \label{req:RQ3.3}  
    The system must categorize recommended actions (e.g., “Immediate Patch,” “Monitor and Prepare Patch,” “Deprioritize”).
\end{itemize}

\subsection{RQ4: A Model and an Algorithm to Rank-Order All Known Classified Vulnerabilities}
\label{req:RQ4}
Finally, the system should prioritize vulnerabilities according to their severity and data completeness.

\begin{itemize}
    \item \textbf{RQ4.1: Severity-Based Ranking}\\
    \label{req:RQ4.1}  
    The system must provide a ranked list of all classified vulnerabilities in descending order of severity.

    \item \textbf{RQ4.2: Priority for Incomplete Data}\\
    \label{req:RQ4.2}  
    Vulnerabilities with missing data must be placed at the top of the list to prompt immediate investigation.

    \item \textbf{RQ4.3: Continuous Updating}\\
    \label{req:RQ4.3}  
    The system must recalculate the ranking daily to account for new information or changing conditions.
\end{itemize}

\section{Non-Functional Requirements}
Non-functional requirements capture the overall qualities and constraints of the system, ensuring it remains usable, scalable, maintainable, and reliable over time.

\begin{itemize}
    \item \textbf{RQNF1: Usability}\\
    \label{req:nonf1}  
    The user interface must be intuitive and provide clear explanations for scores, remediation options, and any missing data.
    
    \item \textbf{RQNF2: Scalability}\\
    \label{req:nonf2}  
    The system must efficiently handle large volumes of vulnerabilities to ensure timely classification and updates.
    
    \item \textbf{RQNF3: Maintainability}\\
    \label{req:nonf3}  
    Configuration for scoring parameters must be easily adjustable without requiring changes to the core system logic.
    
    \item \textbf{RQNF4: Reliability}\\
    \label{req:nonf4}  
    Data retrieval from external sources must be robust, with appropriate failure handling to avoid partial or corrupted information.
\end{itemize}